The evolution of gene expression in the term placenta of viviparous mammals

### 2015-08-19 manuscript/manuscript.Rnw
site mossman and carter

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form (Mossman,
1937). Species differ in the shape of attachment (i.e., discoid,
diffuse, cotyledonary, zonary), degree of invasiveness of fetal
tissues into maternal tissues (i.e. epitheliochorial,
endotheliochorial, hemochorial), and degree of interdigitation of
maternal-fetal tissue (i.e., villous, folded, trabecular, lamellar,
labrynthine) (Mossman, 1987). In addition, therian mammals differ
drastically in the number and size of their offspring, as well as the
length of gestation, traits often associated with specific aspects of
placental morphology (Capellini et al., 2011).

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental”
mammals. The term “placental mammals” is misleading, as marsupials
also possess a placenta, although its direct apposition to maternal
tissues lasts for a proportionally shorter time during gestation
(Renfree, 2010). Marsupials also differ from eutherian mammals in
that most possess only a choriovitelline placenta, produced from the
fusion of the chorion and yolk sac, during most of gestation. Although
some eutherian mammals may possess a choriovitelline placenta at some
point in development, the major site of nutrient exchange in
placentals occurs via the chorioallantoic placenta, produced from the
fusion of the chorion and allantois (Mossman, 1987). It is unclear
whether the first placenta was more like that of marsupials or
eutherians; however, using phylogenetic methods, the ancestral
eutherian was reconstructed as possessing a relatively invasive
hemochorial placenta, in which the maternal and fetal blood supply is
separated by one tissue layer, as in humans and mouse (Elliot
and Crespi, 2009; Wildman et al., 2006).

Although the morphology of the mammalian placenta has been well
characterized in a number of species (Mossman, 1987), the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective (Carter, 2012). Studies of certain model species
(i.e., humans, mice, cattle) have revealed a stunning divergence in
the genetics of the placenta, especially in later stages of pregnancy
(Carter, 2012; Hou et al., 2012; Knox and Baker, 2008; Su et al.,
2002). For example, each lineage has evolved expansive
placenta-specific gene families with expression patterns that can vary
in space and time (Green et al., 2000; Simmons et al., 2008). These
include independent expansions of the prolactins in mouse and cow,
pregnancy-specific glycoproteins (PSGs) in humans and mouse,
pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
the growth hormone and chorionic gonadotropin beta genes in hominoid
primates (Haig, 2008; Rawn and Cross, 2008). Expression patterns of
specific genes also can vary widely between species. For example, the
expression of orthologous genes in the placenta have been found to
vary substantially between humans and mice (Su et al., 2002), and the
recruitment of SIGLEC6 to placental expression in humans differs from
both chimpanzees and mice (Brinkman-Van der Linden et al., 2007).

>>>>

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology (Capellini et al., 2011).

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental”
mammals. The term “placental mammals” is misleading, as marsupials
also possess a placenta, although its direct apposition to maternal
tissues lasts for a proportionally shorter time during gestation
(Renfree, 2010). Marsupials also differ from eutherian mammals in
that most possess only a choriovitelline placenta, produced from the
fusion of the chorion and yolk sac, during most of gestation. Although
some eutherian mammals may possess a choriovitelline placenta at some
point in development, the major site of nutrient exchange in
placentals occurs via the chorioallantoic placenta, produced from the
fusion of the chorion and allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
whether the first placenta was more like that of marsupials or
eutherians; however, using phylogenetic methods, the ancestral
eutherian was reconstructed as possessing a relatively invasive
hemochorial placenta, in which the maternal and fetal blood supply is
separated by one tissue layer, as in humans and mouse (Elliot
and Crespi, 2009; Wildman et al., 2006).

Although the morphology of the mammalian placenta has been well
characterized in a number of species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}. Studies of certain model species
(i.e., humans, mice, cattle) have revealed a stunning divergence in
the genetics of the placenta, especially in later stages of pregnancy
(Carter, 2012; Hou et al., 2012; Knox and Baker, 2008; Su et al.,
2002). For example, each lineage has evolved expansive
placenta-specific gene families with expression patterns that can vary
in space and time (Green et al., 2000; Simmons et al., 2008). These
include independent expansions of the prolactins in mouse and cow,
pregnancy-specific glycoproteins (PSGs) in humans and mouse,
pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
the growth hormone and chorionic gonadotropin beta genes in hominoid
primates (Haig, 2008; Rawn and Cross, 2008). Expression patterns of
specific genes also can vary widely between species. For example, the
expression of orthologous genes in the placenta have been found to
vary substantially between humans and mice (Su et al., 2002), and the
recruitment of SIGLEC6 to placental expression in humans differs from
both chimpanzees and mice (Brinkman-Van der Linden et al., 2007).
